---
title: "Project Proposal, Step 2"
subtitle: "EC607"
author: "Due *before* midnight (11:59pm) on .bold[Wednesday, 27 May 2019]"
# date: "Today"
seal: false
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    # self_contained: true
    nature:
      ratio: '8.5:11'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: clear

```{r, setup, include = F}
# Packages
library(pacman)
p_load(
  ggplot2, gridExtra, ggthemes, latex2exp, kableExtra,
  tidyverse, broom, knitr, magrittr
)
# Colors
red_pink <- "#e64173"
turquoise <- "#20B2AA"
grey_light <- "grey70"
grey_mid <- "grey50"
grey_dark <- "grey20"
# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  warning = F,
  message = F
)
```

.mono.b[DUE:] Your .b[typed] proposal is due *before* 11:59pm on Wednesday, 27 May 2019 on [Canvas](https://canvas.uoregon.edu/).

Your grade will depend upon your ability to clearly respond to the assignment—concepts *and* constraints.

.mono.b[Assignment] Submit a project proposal that

1\. Motivates and outlines your .hi[*causal* question] of interest.
<br>
2\. Explains .hi[potential sources of selection] that could lead to bias.
<br>
3\. Describes the .hi[ideal experiment] which one could use to answer this question
<br>
4\. Discusses a .hi[practical research design] through which one could answer the original question—clarifying *how* this research design avoids selection bias.

.hi[Note:] Your question **must be causal in nature**. If it is not, come up with a new question.

.mono.b[Sections:] Your proposal requires the following sections/components. **Separate them.**
<br>After each section, state the section's word count (*e.g.*, *13 words*).

.mono.b[Title:] (1 ≤ words ≤ 15)
<br>A title that clearly describes your question—and potentially how you would answer it.

.mono.b[Abstract:] (50 ≤ words ≤ 150)
<br>A brief description of your project. Clearly describe the main question, how you will answer it, and why/for whom the results matter. Be concise and clear. You hook the reader here and elaborate later.

.mono.b[Question and motivation:] (100 ≤ words ≤ 300)
<br> Explain why this area of research is interesting/important in general (not just to you). Why should your reader care/keep reading? After you .b[briefly] motivate the general topic, clearly describe your specific .b[causal question]. If necessary, motivate the specific question too.

.mono.b[Selection:] (100 ≤ words ≤ 300)
<br> Why is this question challenging to answer empirically? In other words, what sources of selection bias concern you? If we simply run a regression of $y$ on $X$, why might $\hat{\beta}$ be biased?

.mono.b[Ideal experiment:] (100 ≤ words ≤ 300)
<br> Describe the ideal experiment that would answer your question. This *ideal* experiment does not need to be practical—*i.e.*, you do not need to be able to run it in real life.

.mono.b[Practical research design:] (100 ≤ words ≤ 400)
<br> How might you causally answer the your question in *real life*? Which **data** would you need? What sort of **research design** would you apply—selection on observables (regression with many controls, matching, propensity-score methods, *etc.*) or selection on observables (IV, RD, *etc.*)? How does this proposed research design **avoid selection bias**?

.mono.b[Update:] Now that you've worked through these components, rewrite your abstract and title.

```{r, generate pdfs, include = F, eval = F}
pagedown::chrome_print("project-002.html", output = "project-002.pdf")
```